---
title: "EDA-challenge.Rmd"
output: html_document
date: "2024-02-17"
---
Keerthi Stanley


Load data into r
```{r}
library(tidyverse)
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/data-wrangling.csv"
d <- read.table(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
```

1. create a new variable, BSD, which is the ratio of average male to female body mass
```{r}
d$BSD <- d$Body_mass_male_mean / d$Body_mass_female_mean
```

2. create a new variable named sex_ratio, which is the ratio of the number of adult females to adult males in a typical group
```{r}
d$sex_ratio <- d$AdultFemale / d$AdultMales
```

3. create a new variable named DI (for “defensibility index”), which is the ratio of day range length to the diameter of the home range

diameter of homerange:

- presume homerange is the circle's area, what is the diameter?

AREA = A = pi*(r^2)
. The variable pi (note, without parentheses) will return the value of 
 as a constant built-in to R, and the function sqrt() can be used to calculate the square root ()
 
diameter = 2*r
 r = sqrt(Area/pi) = sqrt(HomeRange_km2/pi)
 
thus the diameter of homerange =
  2*r = 2*(sqrt(HomeRange_km2/pi))
  
but remember to draw HomeRange_km2 from the dataset using d$
```{r}
d$DI <- d$DayLength_km / (2 * sqrt(d$HomeRange_km2/pi))
```
________________________________________
4. Plot the relationship between day range length and time spent moving, for these primate species overall and by family. 

In the table we see that:
DayLength_km = day range length (y)
Move = time spent moving (x)


without any transformations:
```{r}
# plot object and color points by Family
p <- ggplot(data = d, aes(x = Move, y = DayLength_km, 
    color = factor(Family))) 

# axis labels
p <- p + xlab("Time Spent Moving") + ylab("Day Range Length (km)")
# make a scatterplot, omit missing data with na.rm
p <- p + geom_point(na.rm = TRUE)
# modify legend
p <- p + theme(legend.position = "bottom", legend.title = element_blank())

p

```
the plot above WIHTOUT a transformation does not give us the best representation of the data, the points are too clustered in one area. so we will replot with a Log transformation

Log transformation:
```{r}
# plot object and color points by Family
p <- ggplot(data = d, aes(x = log(Move), y = log(DayLength_km),
    color = factor(Family)))

# axis labels
p <- p + xlab("log(Time Spent Moving)") + ylab("log(Day Range Length (km))")
# make a scatterplot, omit missing data with na.rm
p <- p + geom_point(na.rm = TRUE)
# modify legend
p <- p + theme(legend.position = "bottom", legend.title = element_blank())

p

```

QUESTIONS:
Do species that spend more time moving travel farther overall?
  - no, the scatter plot does not show a clear positive linear slope that would indicate that species that spend more time moving also travel further
   
How about within any particular primate family? 
  - no even within each family there is not a clear trend correlating more time moving to farther distance traveled
  
Should you transform either of these variables?
  - yes, as shown above, the log transformation of x and y gives us a much more normal data distribution

_________________________________________________
5. Plot the relationship between day range length and group size, overall and by family.


WITHOUT TRANSFORMATION:
```{r}
# plot object and color points by Family
p <- ggplot(data = d, aes(x = MeanGroupSize, y = DayLength_km,
    color = factor(Family)))

# axis labels
p <- p + xlab("Group Size") + ylab("Day Range Length (km)") 
# make a scatterplot, omit missing data with na.rm
p <- p + geom_point(na.rm = TRUE)
# modify legend
p <- p + theme(legend.position = "bottom", legend.title = element_blank())

p
```

Log Transformation:
```{r}
p <- ggplot(data = d, aes(x = log(MeanGroupSize), y = log(DayLength_km), 
    color = factor(Family)))  # plot object and color points by Family

# axis labels
p <- p + xlab("log(Group Size)") + ylab("log(Day Range Length (km))")
# make a scatterplot, omit missing data with na.rm
p <- p + geom_point(na.rm = TRUE)
# modify legend
p <- p + theme(legend.position = "bottom", legend.title = element_blank())

p
```

QUESTIONS:
Do species that live in larger groups travel farther overall? 
  - certain families do show this trend, with a positive linear slope in their data; however, not all of the groups show this so we cannot assume that this is an overall trend
  
How about within any particular primate family? 
  - some families, such as Cercopithecidae, Hominidae, and Lorisidae do suggest that when they live in larger groups, they also travel farther; other groups, like Lemuridae, show no significant correlation between the two variables
  
Should you transform either of these variables?
  - Once again the log transformation of both variables allows us to best visualize the data
  
_________________________________________________
6. Plot the relationship between body size dimorphism and canine size dimorphism overall and by family.

BSD = body size dimorphism
Canine_Dimorphism = canine dimorphism

using the log transformation for this as well, as it best visualizes the data:
```{r}
p <- ggplot(data = d, aes(x = log(BSD), y = log(Canine_Dimorphism),
    color = factor(Family)))  # plot object and color points by Family

# modify the axis labels
p <- p + xlab("log(Body Size Dimorphism)") + ylab("log(Canine Dimorphism)")
# then we make a scatterplot, omit missing data with na.rm
p <- p + geom_point(na.rm = TRUE)
# then we modify the legend
p <- p + theme(legend.position = "bottom", legend.title = element_blank())
# and, finally, we plot the object
p
```
QUESTIONS:
Do taxa with greater size dimorphism also show greater canine dimorphism?
  - yes! we do tend to see a more positive linear trend here OVERALL throughout families
  
___________________________________________________
7. Create a new variable named diet_strategy that is “frugivore” if fruits make up >50% of the diet, “folivore” if leaves make up >50% of the diet, and “omnnivore” if neither of these is true. Then, do boxplots of group size for species with different dietary strategies. 

1st- create new variable diet_strategy
    --> use mutate()
    
2nd- use conditional expression
    --> in this case case_when() because we are dealing with multiple "if" statements
    
- here I'm using the pipe operator to join the two functions
```{r}
# modify dataset d
# mutate()- adds column titled 'diet_strategy'
d <- d %>%
  mutate(diet_strategy = case_when(
    Fruit > 50 ~ "frugivore",
    Leaves > 50 ~ "folivore",
    TRUE ~ "omnivore" #basically applies to all data that doesn't fit into the original 2 criteria/cases mentioned
  )
  )
```
Now time to make the BOXPLOTS for this data (x= species with same diet strategy, y= group size):
```{r}
p <- ggplot(data = d, aes(x = diet_strategy, y = log(MeanGroupSize)))
p <- p + geom_boxplot(na.rm = TRUE) #plot boxplot and omit missing data
p <- p + theme(axis.text.x = element_text(angle = 90))
p <- p + ylab("log(Mean Group Size)")
p <- p + xlab("Diet Strategy")
p
```

QUESTIONS:
Do frugivores live in larger groups than folivores?
  - no, while the median of mean group sizes for frugivores is slightly higher than that of folivores, the interquartile ranges of both overlap so much that this is not a significant difference

------------------------
8. In one line of code, using {dplyr} verbs and the forward pipe (%>%) operator, do the following:

  a) Add a variable, Binomial to the data frame d, which is a concatenation of the Genus and Species…
- same process as before with mutate()
  
  b) Trim the data frame to only include the variables Binomial, Family, Brain_size_species_mean, and Body_mass_male_mean
- use select() from dplyr

  c) Group these variables by Family
- use group_by() from dplyr

  d) Calculate the average value for Brain_size_species_mean and Body_mass_male_mean per Family (remember, you may need to specify na.rm = TRUE)
- use summarize()
  
  e) arrange by increasing average brain size
- use arrange()
  
```{r}
# the isntructions said to make these changes to dataset d, but I didn't want to alter the original changes to the dataset made from the previous problems, so I created the new dataset, d2

# if you want to directly change d, then change d2 to d below
d2 <- d %>%
  # first let's add/mutate in the new variable Binomial
  mutate(Binomial = paste(Genus, Species, sep = " ")) %>%
  # select only the variables we want to still include (listed above)
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) %>%
  
  # NOTE!! - if you want "Binomial" to be represented in the dataset then stop at line 224 and remove the pipe operator!
  
  # group by Family from dataset d
  group_by(Family) %>%
  #calculate averages for brain size and male body mass
  summarize(avgBrainSize = mean(Brain_Size_Species_Mean, na.rm = TRUE),
    avgMale_BodyMass = mean(Body_mass_male_mean, na.rm = TRUE)) %>%
  #arrange by increasing average brain size
  arrange(avgBrainSize)
```
  
  
  
  
  
  
  
  
